---
title: "Final Project"
author: "Group Rimundo Lulio"
date: "2023-12-14"
output:
  prettydoc::html_pretty:
    theme: cayman
    highlight: github
---

This report presents the findings and conclusions of the final project for the Data Mining course. The project focused on analyzing a dataset provided by the CDC, containing information from the Behavioral Risk Factor Surveillance System (BRFSS). The BRFSS conducts annual telephone surveys to gather data on the health status of U.S. residents, the dataset used in this project comprises information on 445,132 residents.

The initial presentation of this project covered the data cleaning process in detail. As such, this report will not delve into the data cleaning procedures but will instead highlight the initial questions posed and the subsequent analyses conducted to address them.


We start by loading and taking a look at the data.

```{r}
df <- read.csv("FPdata.csv")
str(df)
```
As we can clearly see we encountered our first problem. “FPdata.csv” has bad formatting, so we use RegEx to “clean” it. 

```{r}
# Read all the lines of the file
lineas <- readLines('FPdata.csv')

# Remove start and end quotes
lineas <- gsub("^\"", "", lineas)
lineas <- gsub("\"$", "", lineas)
# Change doble quotes to single ones
lineas <- gsub("\"\"", "\"", lineas)

# Create the new df from a tmep file
archivo_temporal <- tempfile(fileext = ".csv")
writeLines(lineas, archivo_temporal)

# Read the cleaned CSV
df <- read.csv(file = archivo_temporal, header = TRUE, stringsAsFactors = TRUE, quote = "\"")

#To save time in the future we’ll save this to a new file called “FPdata_CLEAN.csv”.
#write.csv(df, "FPdata_CLEAN.csv", row.names = FALS
```

As we can observe above the inital project with the corrupted data has been addressed, now we begin to do the data cleaning itself, first of all knowing that this is a telephone survey we will check entries that have more than five nan's, as this could mean that the call has been interrupted and the data could not be 'real'.

Let’s define a threshold for NA’s per show and visualize that rows.

```{r}
threshold <- 5
rows_with_many_nas <- which(rowSums(is.na(df)) > threshold)
df[rows_with_many_nas, ]
df <- df[-rows_with_many_nas, ]
```

As we can see we have 40 values that follows our criteria, compared to the aproximate 400k entries of the dataset this could be removed without a major impact.

To not mess with the original data, we create a “copy” of it and manipulate it. Usually we transform factors to int assigning an ascendent number following alphabetical order. In other cases we binarize the columns, with the first level being 0. In some cases we binarize columns with more than two levels to simplify, take RemovedTeeth as exemple, where 0 means “None of them” and 1 any of the rest. If keeping the NA’s makes any sense, -1 is assigned to that values.

```{r}
cdf <- data.frame(df)
```

Let's take a look column by columns and we will define how we will treat them.

## State

State is a categorical value, our dataset need to be transformed to numerical data inorder to be used by the algorithms seen during the course, to code this column we will transform each factor to a int that represent each state.

```{r}
table(df$State)
cdf$State <- factor(df$State)

# Asign a unique value to each factor
cdf$State <- as.integer(cdf$State)
```

## Gender

Gender, in this case, is a binary class male or female, so we will do a binary encoding where 0 represents female and 1 male.

```{r}
table(df$Gender, useNA = "ifany")

cdf$Gender <- ifelse(df$Gender == "Female", 0, 1)
table(cdf$Gender)
```

## General health

General health is the same type as state but we have nan's, to deal with them we will initialy fill them with the mode, that in this case is 4.

```{r}
numeric_general_health <- factor(df$GeneralHealth, levels = c("Poor", "Fair", "Good", "Very good", "Excellent"))

#Fill NAN's with mode (4)
cdf$GeneralHealth <- ifelse(is.na(numeric_general_health), 4, as.integer(numeric_general_health))
table(cdf$GeneralHealth)
```

## PhysicalHealthDays

PhysicalHealthDays is a numeric value that has nan's, because it won't make sense to have 3.2 days off or 0.4 we won't calculate the mean and we will use the mode again.

```{r}
PhysicalHealthDays_frq <- table(df$PhysicalHealthDays, useNA = "ifany")
mode_PhysicalHealthDays <- names(PhysicalHealthDays_frq[which.max(PhysicalHealthDays_frq)])

cdf$PhysicalHealthDays <- ifelse(is.na(df$PhysicalHealthDays), as.numeric(mode_PhysicalHealthDays),as.numeric(df$PhysicalHealthDays))

table(cdf$PhysicalHealthDays)
```
## MentalHealthDays

We will procede as same as with PhysicalHealthDays.

```{r}
MentalHealthDays_frq <- table(df$MentalHealthDays, useNA = "ifany") 
mode_MentalHealthDays <- names(MentalHealthDays_frq[which.max(MentalHealthDays_frq)])

cdf$MentalHealthDays <- ifelse(is.na(df$MentalHealthDays), as.numeric(mode_MentalHealthDays),as.numeric(df$MentalHealthDays))

table(cdf$MentalHealthDays)
```

## LastCheckupTime

With this column we will convert each categorical factor to a integer representing that class, and a -1 when we find any NAN, in this case we don't have any.

```{r}
table(df$LastCheckupTime, useNA = "ifany") 

numeric_checkup_time <- factor(df$LastCheckupTime)

cdf$LastCheckupTime <- ifelse(is.na(numeric_checkup_time), -1, as.integer(numeric_checkup_time))
table(cdf$LastCheckupTime)
```

## PhysicalActivities

PhysicalActivities is a binary class with some columns being blank saces, representing a yes, and orinary yes and no.

```{r}
table(df$PhysicalActivities, useNA = "ifany") 
cdf$PhysicalActivities <- ifelse(df$PhysicalActivities == "", 1, ifelse(df$PhysicalActivities == "Yes", 1,0))
table(cdf$PhysicalActivities)
```

## SleepHours

SleepHours we will fill with the mode any nan, later we will normalize this continuous values between 0 and 1.

```{r}
SleepHours_frq <- table(df$SleepHours, useNA = "ifany") 
mode_SleepHours <- names(SleepHours_frq[which.max(SleepHours_frq)])

cdf$SleepHours <- ifelse(is.na(df$SleepHours), as.numeric(mode_SleepHours),as.numeric(df$SleepHours))
table(cdf$SleepHours)
```
## RemovedTeeth

RemovedTeeth is as same as LastCheckupTime

```{r}
table(df$RemovedTeeth, useNA = "ifany")
cdf$RemovedTeeth <- ifelse(df$RemovedTeeth == "", 1, ifelse(grepl("None", df$RemovedTeeth),0,1))
table(cdf$RemovedTeeth)
```

## HadHeartAttack

Hearth atack is as same as PhysicalActivities

```{r}
table(df$HadHeartAttack, useNA = "ifany") 
cdf$HadHeartAttack <- ifelse(df$HadHeartAttack == "", 0, ifelse(df$HadHeartAttack == "No", 0,1))
table(cdf$HadHeartAttack)
```

## HadAngina

HadAngina is as same as PhysicalActivities

```{r}
table(df$HadAngina, useNA = "ifany") 
cdf$HadAngina <- ifelse(df$HadAngina == "", 0, ifelse(df$HadAngina == "No", 0,1))
table(cdf$HadAngina)
```

## HadStroke

Same as PhysicalActivities

```{r}
table(df$HadStroke, useNA = "ifany") 
cdf$HadStroke <- ifelse(df$HadStroke == "", 0, ifelse(df$HadStroke == "No", 0,1))
table(cdf$HadStroke)
```
## HadAsthma

Same as PhysicalActivities

```{r}
table(df$HadAsthma, useNA = "ifany") 
cdf$HadAsthma <- ifelse(df$HadAsthma == "", 0, ifelse(df$HadAsthma == "No", 0,1))
table(cdf$HadAsthma)
```
## HadSkinCancer

Same as PhysicalActivities

```{r}
table(df$HadSkinCancer, useNA = "ifany") 
cdf$HadSkinCancer <- ifelse(df$HadSkinCancer == "", 0, ifelse(df$HadSkinCancer == "No", 0,1))
table(cdf$HadSkinCancer)
```

## HadCOPD

Same as PhysicalActivities

```{r}
table(df$HadCOPD, useNA = "ifany") 
cdf$HadCOPD <- ifelse(df$HadCOPD == "", 0, ifelse(df$HadCOPD == "No", 0,1))
table(cdf$HadCOPD)
```

## HadDepressiveDisorder

Same as PhysicalActivities

```{r}
table(df$HadDepressiveDisorder, useNA = "ifany") 
cdf$HadDepressiveDisorder <- ifelse(df$HadDepressiveDisorder == "", 0, ifelse(df$HadDepressiveDisorder == "No", 0,1))
table(cdf$HadDepressiveDisorder)
```

## HadKidneyDisease

Same as PhysicalActivities

```{r}
table(df$HadKidneyDisease, useNA = "ifany") 
cdf$HadKidneyDisease <- ifelse(df$HadKidneyDisease == "", 0, ifelse(df$HadKidneyDisease == "No", 0,1))
table(cdf$HadKidneyDisease)
```

## HadArthritis

Same as PhysicalActivities

```{r}
table(df$HadArthritis, useNA = "ifany") 
cdf$HadArthritis <- ifelse(df$HadArthritis == "", 0, ifelse(df$HadArthritis == "No", 0,1))
table(cdf$HadArthritis)
```

## HadDiabetes

HAdDiaetes is a special case, where we have 3 values, 'No', 'No, pre-diabetes or borderline diabetes' and yes, our group opted to binarize this columns with yes and no, grouping the first two group at one.

```{r}
table(df$HadDiabetes, useNA = "ifany") 
cdf$HadDiabetes <- ifelse(df$HadDiabetes == "", 0, ifelse(grepl("No",df$HadDiabetes), 0,1))
table(cdf$HadDiabetes)
```

## DeafOrHardOfHearing

Same as PhysicalActivities

```{r}
table(df$DeafOrHardOfHearing, useNA = "ifany") 
cdf$DeafOrHardOfHearing <- ifelse(df$DeafOrHardOfHearing == "", 0, ifelse(df$DeafOrHardOfHearing == "No", 0,1))
table(cdf$DeafOrHardOfHearing)
```

## BlindOrVisionDifficulty

Same as PhysicalActivities

```{r}
table(df$BlindOrVisionDifficulty, useNA = "ifany") 
cdf$BlindOrVisionDifficulty <- ifelse(df$BlindOrVisionDifficulty == "", 0, ifelse(df$BlindOrVisionDifficulty == "No", 0,1))
table(cdf$BlindOrVisionDifficulty)
```

## DifficultyConcentrating

Same as PhysicalActivities

```{r}
table(df$DifficultyConcentrating, useNA = "ifany") 
cdf$DifficultyConcentrating <- ifelse(df$DifficultyConcentrating == "", 0, ifelse(df$DifficultyConcentrating == "No", 0,1))
table(cdf$DifficultyConcentrating)
```
## DifficultyWalking

Same as PhysicalActivities

```{r}
table(df$DifficultyWalking, useNA = "ifany") 
cdf$DifficultyWalking <- ifelse(df$DifficultyWalking == "", 0, ifelse(df$DifficultyWalking == "No", 0,1))
table(cdf$DifficultyWalking)
```
## DifficultyDressingBathing

Same as PhysicalActivities

```{r}
table(df$DifficultyDressingBathing, useNA = "ifany") 
cdf$DifficultyDressingBathing <- ifelse(df$DifficultyDressingBathing == "", 0, ifelse(df$DifficultyDressingBathing == "No", 0,1))
table(cdf$DifficultyDressingBathing)
```

## DifficultyErrands

Same as PhysicalActivities

```{r}
table(df$DifficultyErrands, useNA = "ifany") 
cdf$DifficultyErrands <- ifelse(df$DifficultyErrands == "", 0, ifelse(df$DifficultyErrands == "No", 0,1))
table(cdf$DifficultyErrands)
```

## SmokerStatus

This case is similar to HadDiabetes, where we only care if they smoke or not, we will binarice this column.

```{r}
table(df$SmokerStatus, useNA = "ifany") 
cdf$SmokerStatus <- ifelse(df$SmokerStatus == "", 0, ifelse(grepl("Never smoked",df$SmokerStatus), 0,1))
table(cdf$SmokerStatus)
```

## ECigarretteUsage

Same as SmokingStatus, grouping by no and yes, at the moment of the survey.

```{r}
table(df$ECigaretteUsage, useNA = "ifany") 
cdf$ECigaretteUsage <- ifelse(df$ECigaretteUsage == "", 0, ifelse(grepl("Never smoked",df$ECigaretteUsage) | grepl("Not at all",df$ECigaretteUsage) , 0,1))
table(cdf$ECigaretteUsage)
```

## ChestScan

Same as PhysicalActivities

```{r}
table(df$ChestScan, useNA = "ifany") 
cdf$ChestScan <- ifelse(df$ChestScan == "", 0, ifelse(df$ChestScan == "No", 0,1))
table(cdf$ChestScan)
```


## RaceEthnicityCategory

Same as LastCheckupTime

```{r}
cdf$RaceEthnicityCategory <- df$RaceEthnicityCategory
table(df$RaceEthnicityCategory, useNA = "ifany") 

cdf$RaceEthnicityCategory[df$RaceEthnicityCategory == ""] <- "White only, Non-Hispanic"
cdf$RaceEthnicityCategory <- factor(cdf$RaceEthnicityCategory)
  cdf$RaceEthnicityCategory <- as.integer(cdf$RaceEthnicityCategory)
  table(cdf$RaceEthnicityCategory)
```

## AgeCategory

Same as LastCheckupTime

```{r}
table(df$AgeCategory, useNA = "ifany") 
cdf$AgeCategory[df$AgeCategory == ""] <- "Age 65 to 69"
cdf$AgeCategory <- factor(cdf$AgeCategory)
  cdf$AgeCategory <- as.integer(cdf$AgeCategory)
  table(cdf$AgeCategory)
```

## HeightInMeters

Same as PhysicalHealthDays

```{r}
height_frq <- table(df$HeightInMeters, useNA = "ifany") 

mode_height <- names(height_frq[which.max(height_frq)])
cdf$HeightInMeters <- ifelse(is.na(df$HeightInMeters),as.numeric(mode_height),as.numeric(df$HeightInMeters) )

table(cdf$HeightInMeters, useNA = "ifany") 
```
## WeightInKilograms

Same as PhysicalHealthDays

```{r}
weight_frq <- table(df$WeightInKilograms) 

mode_weight <- names(weight_frq[which.max(weight_frq)])
cdf$WeightInKilograms <- ifelse(is.na(df$WeightInKilograms),as.numeric(mode_weight),as.numeric(df$WeightInKilograms) )

table(cdf$WeightInKilograms, useNA = "ifany") 
```
## BMI

As we know BMI is a linear combinarion of Height and Weigth, first of all we will take a look at the distribution of the data to check if we observe some strange data such as outliers.

```{r}
BMI_frq <- table(df$BMI) 
mode_BMI <- names(BMI_frq[which.max(BMI_frq)])
cdf$BMI <- ifelse(is.na(df$BMI),as.numeric(mode_BMI),as.numeric(df$BMI) )
hist(cdf$BMI)
```

As we can see this doesn't follow a normal distributon and presents outliers, we will calculate the IQR to check which indidividuals exceed the 1.5 IQR. 

```{r}
quantiles = quantile(cdf$BMI)
q1 <- quantiles[2]
q3 <- quantiles[4]

iqr = q3 - q1
lower_bound = q1 - 1.5 * iqr
upper_bound = q3 + 1.5 * iqr

outliers <- cdf$BMI[cdf$BMI < lower_bound | cdf$BMI > upper_bound]
cdf[cdf$BMI < lower_bound | cdf$BMI > upper_bound, ]
df <- cdf[!(cdf$BMI < lower_bound | cdf$BMI > upper_bound), ]
cdf <- cdf[!(cdf$BMI < lower_bound | cdf$BMI > upper_bound), ]
```

Now that we have deleted the outliers, arround 20k of them, we will take a look again at the distribution.

```{r}
hist(cdf$BMI)
```

New can observe that we didn't improve much our distribution, upon further research we can observe that we have quite larges BMI measure, at the begging this socked us a bit but then we remembered that this are entries from US residents and we know that they tend to have overweight problems so we will no treat this values as outliers and will be used as they are.

## AlcoholDrinkers

Same as PhysicalActivities

```{r}

table(cdf$AlcoholDrinkers, useNA = "ifany") 
cdf$AlcoholDrinkers <- ifelse(cdf$AlcoholDrinkers == "", 1, ifelse(cdf$AlcoholDrinkers == "No", 0,1))
table(cdf$AlcoholDrinkers)
```

## HIVTesting

Same as PhysicalActivities

```{r}
table(cdf$HIVTesting, useNA = "ifany") 
cdf$HIVTesting <- ifelse(cdf$HIVTesting == "", 0, ifelse(cdf$HIVTesting == "No", 0,1))
table(cdf$HIVTesting)
```

## FluVaxLast12

Same as PhysicalActivities

```{r}
table(cdf$FluVaxLast12, useNA = "ifany") 
cdf$FluVaxLast12 <- ifelse(cdf$FluVaxLast12 == "", 1, ifelse(cdf$FluVaxLast12 == "No", 0,1))
table(cdf$FluVaxLast12)
```

## PneumoVaxEver

Same as PhysicalActivities

```{r}
table(cdf$PneumoVaxEver, useNA = "ifany") 
cdf$PneumoVaxEver <- ifelse(cdf$PneumoVaxEver == "", 0, ifelse(cdf$PneumoVaxEver == "No", 0,1))
table(cdf$PneumoVaxEver)
```

## TetanusLast10Tdap

Same as HadDiabetes or SmokerStatus, we binarize this column.

```{r}
tetanus_frq <- table(cdf$TetanusLast10Tdap, useNA = "ifany") 
tetanus_frq
mode_TetanusLast10Tdap <- names(tetanus_frq[which.max(tetanus_frq)])
mode_TetanusLast10Tdap
cdf$TetanusLast10Tdap <- ifelse(cdf$TetanusLast10Tdap == "",0,ifelse(grepl(cdf$TetanusLast10Tdap, "No"), 0, 1) )
table(cdf$TetanusLast10Tdap, useNA = "ifany")
```

## HighRiskLastYear

Same as PhysicalActivities

```{r}
table(cdf$HighRiskLastYear, useNA = "ifany") 
cdf$HighRiskLastYear <- ifelse(cdf$HighRiskLastYear == "", 0, ifelse(cdf$HighRiskLastYear == "No", 0,1))
table(cdf$HighRiskLastYear)
```

## CovidPos

Same as PhysicalActivities

```{r}
cdf$CovidPos <- df$CovidPos
table(df$CovidPos, useNA = "ifany") 
cdf$CovidPos <- ifelse(df$CovidPos == "",0,ifelse(grepl("No", df$CovidPos), 0, 1))
table(cdf$CovidPos, useNA = "ifany")
```

Finally we have all of our data cleaned and addressed problems such as NAN's or outliers, to check we will run a str command to be sure.

```{r}
str(cdf)
```

Finally we will look at the correlation matrix to take a look at the data, see if there's some correlated columns that surprises us and helps with the formulation of the questions.

```{r}
heatmap(cor(cdf))
```

```{r}
write.csv(df, "FPdata_CLEAN.csv", row.names = FALSE)
```


