---
title: "Final Project - Macia - 2 - Data Preparation"
author: "Josep Damià Ruiz Pons and Maçià Bartomeu Llabrés Ferrer"
date: "2023-12-14"
output: html_document
---

We start by loading the previous cleaned file and check if the data is the same.
```{r}
df <- read.csv("FPdata_CLEAN.csv", stringsAsFactors = TRUE)
str(df)
summary(df)
```

Let's define a threshold for NA's per show and visualize that rows.
```{r}
threshold <- 5
rows_with_many_nas <- which(rowSums(is.na(df)) > threshold)
df[rows_with_many_nas, ]
df <- df[-rows_with_many_nas, ]
```

To not mess with the original data, we create a "copy" of it and manipulate it. Usually we transform factors to int assigning an ascendent number following alphabetical order. In other cases we binarize the columns, with the first level being 0. In some cases we binarize columns with more than two levels to simplify, take RemovedTeeth as exemple, where 0 means "None of them" and 1 any of the rest. If keeping the NA's makes any sense, -1 is assigned to that values.
```{r}
# Crear un nuevo data frame con la misma estructura pero sin datos
cdf <- data.frame(df)
```

## State
```{r}
table(df$State)
cdf$State <- factor(df$State)
  # Asigna un número único a cada nivel del factor
  cdf$State <- as.integer(cdf$State)
```

## Gender
```{r}
table(df$Gender, useNA = "ifany")

cdf$Gender <- ifelse(df$Gender == "Female", 0, 1)
table(cdf$Gender)
```

## General health
```{r}
# Convertir el vector de datos a un factor con niveles ordenados de forma ascendente
numeric_general_health <- factor(df$GeneralHealth, levels = c("Poor", "Fair", "Good", "Very good", "Excellent"))
# Asignar -1 a los valores NA utilizando ifelse
cdf$GeneralHealth <- ifelse(is.na(numeric_general_health), 4, as.integer(numeric_general_health))
table(cdf$GeneralHealth)
```

## PhysicalHealthDays
```{r}

PhysicalHealthDays_frq <- table(df$PhysicalHealthDays, useNA = "ifany")
PhysicalHealthDays_frq
mode_PhysicalHealthDays <- names(PhysicalHealthDays_frq[which.max(PhysicalHealthDays_frq)])
mode_PhysicalHealthDays
cdf$PhysicalHealthDays <- ifelse(is.na(df$PhysicalHealthDays), as.numeric(mode_PhysicalHealthDays),as.numeric(df$PhysicalHealthDays))
table(cdf$PhysicalHealthDays)
```

## MentalHealthDays
```{r}

MentalHealthDays_frq <- table(df$MentalHealthDays, useNA = "ifany") 
MentalHealthDays_frq
mode_MentalHealthDays <- names(MentalHealthDays_frq[which.max(MentalHealthDays_frq)])
mode_MentalHealthDays
cdf$MentalHealthDays <- ifelse(is.na(df$MentalHealthDays), as.numeric(mode_MentalHealthDays),as.numeric(df$MentalHealthDays))
table(cdf$MentalHealthDays)
```

## LastCheckupTime
```{r}
table(df$LastCheckupTime, useNA = "ifany") 
# Convertir el vector de datos a un factor con niveles ordenados de forma ascendente
numeric_checkup_time <- factor(df$LastCheckupTime)
# Asignar -1 a los valores NA utilizando ifelse
cdf$LastCheckupTime <- ifelse(is.na(numeric_checkup_time), -1, as.integer(numeric_checkup_time))
table(cdf$LastCheckupTime)
```

## PhysicalActivities
```{r}
table(df$PhysicalActivities, useNA = "ifany") 
cdf$PhysicalActivities <- ifelse(df$PhysicalActivities == "", 1, ifelse(df$PhysicalActivities == "Yes", 1,0))
table(cdf$PhysicalActivities)
```

## SleepHours
```{r}
SleepHours_frq <- table(df$SleepHours, useNA = "ifany") 
SleepHours_frq
mode_SleepHours <- names(SleepHours_frq[which.max(SleepHours_frq)])
mode_SleepHours
cdf$SleepHours <- ifelse(is.na(df$SleepHours), as.numeric(mode_SleepHours),as.numeric(df$SleepHours))
table(cdf$SleepHours)
```

## RemovedTeeth
```{r}
table(df$RemovedTeeth, useNA = "ifany")
cdf$RemovedTeeth <- ifelse(df$RemovedTeeth == "", 1, ifelse(grepl("None", df$RemovedTeeth),0,1))
table(cdf$RemovedTeeth)
```

## HadHeartAttack
```{r}
table(df$HadHeartAttack, useNA = "ifany") 
cdf$HadHeartAttack <- ifelse(df$HadHeartAttack == "", 0, ifelse(df$HadHeartAttack == "No", 0,1))
table(cdf$HadHeartAttack)
```

## HadAngina
```{r}
table(df$HadAngina, useNA = "ifany") 
cdf$HadAngina <- ifelse(df$HadAngina == "", 0, ifelse(df$HadAngina == "No", 0,1))
table(cdf$HadAngina)
```

## HadStroke
```{r}
table(df$HadStroke, useNA = "ifany") 
cdf$HadStroke <- ifelse(df$HadStroke == "", 0, ifelse(df$HadStroke == "No", 0,1))
table(cdf$HadStroke)
```

## HadAsthma
```{r}
table(df$HadAsthma, useNA = "ifany") 
cdf$HadAsthma <- ifelse(df$HadAsthma == "", 0, ifelse(df$HadAsthma == "No", 0,1))
table(cdf$HadAsthma)
```

## HadSkinCancer
```{r}
table(df$HadSkinCancer, useNA = "ifany") 
cdf$HadSkinCancer <- ifelse(df$HadSkinCancer == "", 0, ifelse(df$HadSkinCancer == "No", 0,1))
table(cdf$HadSkinCancer)
```

## HadCOPD
```{r}
table(df$HadCOPD, useNA = "ifany") 
cdf$HadCOPD <- ifelse(df$HadCOPD == "", 0, ifelse(df$HadCOPD == "No", 0,1))
table(cdf$HadCOPD)
```

## HadDepressiveDisorder
```{r}
table(df$HadDepressiveDisorder, useNA = "ifany") 
cdf$HadDepressiveDisorder <- ifelse(df$HadDepressiveDisorder == "", 0, ifelse(df$HadDepressiveDisorder == "No", 0,1))
table(cdf$HadDepressiveDisorder)
```

## HadKidneyDisease
```{r}
table(df$HadKidneyDisease, useNA = "ifany") 
cdf$HadKidneyDisease <- ifelse(df$HadKidneyDisease == "", 0, ifelse(df$HadKidneyDisease == "No", 0,1))
table(cdf$HadKidneyDisease)
```

## HadArthritis
```{r}
table(df$HadArthritis, useNA = "ifany") 
cdf$HadArthritis <- ifelse(df$HadArthritis == "", 0, ifelse(df$HadArthritis == "No", 0,1))
table(cdf$HadArthritis)
```

## HadDiabetes
```{r}
table(df$HadDiabetes, useNA = "ifany") 
cdf$HadDiabetes <- ifelse(df$HadDiabetes == "", 0, ifelse(grepl("No",df$HadDiabetes), 0,1))
table(cdf$HadDiabetes)
```

## DeafOrHardOfHearing
```{r}
table(df$DeafOrHardOfHearing, useNA = "ifany") 
cdf$DeafOrHardOfHearing <- ifelse(df$DeafOrHardOfHearing == "", 0, ifelse(df$DeafOrHardOfHearing == "No", 0,1))
table(cdf$DeafOrHardOfHearing)
```

## BlindOrVisionDifficulty
```{r}
table(df$BlindOrVisionDifficulty, useNA = "ifany") 
cdf$BlindOrVisionDifficulty <- ifelse(df$BlindOrVisionDifficulty == "", 0, ifelse(df$BlindOrVisionDifficulty == "No", 0,1))
table(cdf$BlindOrVisionDifficulty)
```

## DifficultyConcentrating
```{r}
table(df$DifficultyConcentrating, useNA = "ifany") 
cdf$DifficultyConcentrating <- ifelse(df$DifficultyConcentrating == "", 0, ifelse(df$DifficultyConcentrating == "No", 0,1))
table(cdf$DifficultyConcentrating)
```

## DifficultyWalking
```{r}
table(df$DifficultyWalking, useNA = "ifany") 
cdf$DifficultyWalking <- ifelse(df$DifficultyWalking == "", 0, ifelse(df$DifficultyWalking == "No", 0,1))
table(cdf$DifficultyWalking)
```

## DifficultyDressingBathing
```{r}
table(df$DifficultyDressingBathing, useNA = "ifany") 
cdf$DifficultyDressingBathing <- ifelse(df$DifficultyDressingBathing == "", 0, ifelse(df$DifficultyDressingBathing == "No", 0,1))
table(cdf$DifficultyDressingBathing)
```

## DifficultyErrands
```{r}
table(df$DifficultyErrands, useNA = "ifany") 
cdf$DifficultyErrands <- ifelse(df$DifficultyErrands == "", 0, ifelse(df$DifficultyErrands == "No", 0,1))
table(cdf$DifficultyErrands)
```

## SmokerStatus
```{r}
table(df$SmokerStatus, useNA = "ifany") 
cdf$SmokerStatus <- ifelse(df$SmokerStatus == "", 0, ifelse(grepl("Never smoked",df$SmokerStatus), 0,1))
table(cdf$SmokerStatus)
```

## ECigarretteUsage
```{r}
table(df$ECigaretteUsage, useNA = "ifany") 
cdf$ECigaretteUsage <- ifelse(df$ECigaretteUsage == "", 0, ifelse(grepl("Never smoked",df$ECigaretteUsage) | grepl("Not at all",df$ECigaretteUsage) , 0,1))
table(cdf$ECigaretteUsage)
```

## ChestScan
```{r}
table(df$ChestScan, useNA = "ifany") 
cdf$ChestScan <- ifelse(df$ChestScan == "", 0, ifelse(df$ChestScan == "No", 0,1))
table(cdf$ChestScan)
```

## RaceEthnicityCategory
Here are some changes respect the original code from our group. As I try to prove is there is any relation between the race and diseases, is better to keep this column as factor and create a new level for NA's to avoid multicollinearity.
```{r}
cdf$RaceEthnicityCategory <- df$RaceEthnicityCategory
table(df$RaceEthnicityCategory, useNA = "ifany") 

#cdf$RaceEthnicityCategory[df$RaceEthnicityCategory == ""] <- "White only, Non-Hispanic"
levels(cdf$RaceEthnicityCategory)[levels(cdf$RaceEthnicityCategory) == ""] <- "0 - NA's"
cdf$RaceEthnicityCategory <- factor(cdf$RaceEthnicityCategory)
  #cdf$RaceEthnicityCategory <- as.integer(cdf$RaceEthnicityCategory)
  table(cdf$RaceEthnicityCategory)
```

## AgeCategory
```{r}
table(df$AgeCategory, useNA = "ifany") 
cdf$AgeCategory[df$AgeCategory == ""] <- "Age 65 to 69"
cdf$AgeCategory <- factor(cdf$AgeCategory)
  cdf$AgeCategory <- as.integer(cdf$AgeCategory)
  table(cdf$AgeCategory)
```

## HeightInMeters
```{r}
height_frq <- table(df$HeightInMeters, useNA = "ifany") 
height_frq
mode_height <- names(height_frq[which.max(height_frq)])
cdf$HeightInMeters <- ifelse(is.na(df$HeightInMeters),as.numeric(mode_height),as.numeric(df$HeightInMeters) )
table(cdf$HeightInMeters, useNA = "ifany") 
```

## WeightInKilograms
```{r}
weight_frq <- table(df$WeightInKilograms) 
weight_frq
mode_weight <- names(weight_frq[which.max(weight_frq)])
mode_weight
cdf$WeightInKilograms <- ifelse(is.na(df$WeightInKilograms),as.numeric(mode_weight),as.numeric(df$WeightInKilograms) )
table(cdf$WeightInKilograms, useNA = "ifany") 

```

## BMI
```{r}
BMI_frq <- table(df$BMI) 
mode_BMI <- names(BMI_frq[which.max(BMI_frq)])
mode_BMI
cdf$BMI <- ifelse(is.na(df$BMI),as.numeric(mode_BMI),as.numeric(df$BMI) )
hist(cdf$BMI)
```

```{r}
quantiles = quantile(cdf$BMI)
q1 <- quantiles[2]
q3 <- quantiles[4]

iqr = q3 - q1
lower_bound = q1 - 1.5 * iqr
upper_bound = q3 + 1.5 * iqr

outliers <- cdf$BMI[cdf$BMI < lower_bound | cdf$BMI > upper_bound]
cdf[cdf$BMI < lower_bound | cdf$BMI > upper_bound, ]
df <- cdf[!(cdf$BMI < lower_bound | cdf$BMI > upper_bound), ]
cdf <- cdf[!(cdf$BMI < lower_bound | cdf$BMI > upper_bound), ]
```

```{r}
hist(cdf$BMI)
table(cdf$HeightInMeters)
hist(cdf$HeightInMeters)
```

## AlcoholDrinkers
```{r}

table(cdf$AlcoholDrinkers, useNA = "ifany") 
cdf$AlcoholDrinkers <- ifelse(cdf$AlcoholDrinkers == "", 1, ifelse(cdf$AlcoholDrinkers == "No", 0,1))
table(cdf$AlcoholDrinkers)
```

## HIVTesting
```{r}
table(cdf$HIVTesting, useNA = "ifany") 
cdf$HIVTesting <- ifelse(cdf$HIVTesting == "", 0, ifelse(cdf$HIVTesting == "No", 0,1))
table(cdf$HIVTesting)
```

## FluVaxLast12
```{r}
table(cdf$FluVaxLast12, useNA = "ifany") 
cdf$FluVaxLast12 <- ifelse(cdf$FluVaxLast12 == "", 1, ifelse(cdf$FluVaxLast12 == "No", 0,1))
table(cdf$FluVaxLast12)
```

## PneumoVaxEver
```{r}
table(cdf$PneumoVaxEver, useNA = "ifany") 
cdf$PneumoVaxEver <- ifelse(cdf$PneumoVaxEver == "", 0, ifelse(cdf$PneumoVaxEver == "No", 0,1))
table(cdf$PneumoVaxEver)
```

## TetanusLast10Tdap
```{r}
tetanus_frq <- table(cdf$TetanusLast10Tdap, useNA = "ifany") 
tetanus_frq
mode_TetanusLast10Tdap <- names(tetanus_frq[which.max(tetanus_frq)])
mode_TetanusLast10Tdap
cdf$TetanusLast10Tdap <- ifelse(cdf$TetanusLast10Tdap == "",0,ifelse(grepl(cdf$TetanusLast10Tdap, "No"), 0, 1) )
table(cdf$TetanusLast10Tdap, useNA = "ifany")
```

## HighRiskLastYear
```{r}
table(cdf$HighRiskLastYear, useNA = "ifany") 
cdf$HighRiskLastYear <- ifelse(cdf$HighRiskLastYear == "", 0, ifelse(cdf$HighRiskLastYear == "No", 0,1))
table(cdf$HighRiskLastYear)
```

## CovidPos
```{r}
cdf$CovidPos <- df$CovidPos
table(df$CovidPos, useNA = "ifany") 
cdf$CovidPos <- ifelse(df$CovidPos == "",0,ifelse(grepl("No", df$CovidPos), 0, 1))
table(cdf$CovidPos, useNA = "ifany")
```

```{r}
str(cdf)
summary(cdf)
```


As we have factors, now we can not plot the correlation heatmap.
```{r}
#heatmap(cor(cdf))
```

To save time in the future we'll save this to a new file called "FPdata_PREP.csv".
```{r}
write.csv(cdf, "FPdata_PREP.csv", row.names = FALSE)
```